---
title: "class14"
format: html
editor: visual
---

```{r}
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
countData = read.csv(countFile, row.names=1)
head(countData)
```

Q. Complete the code below to remove the troublesome first column from countData

```{r}
countData <- as.matrix(countData[, -1])
head(countData)
```

```{r}
countData = countData[rowSums(countData) > 0, ]
head(countData)
```

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```

```{r}
dds
```

```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
```


Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.


```{r}
res <- results(dds, alpha = 0.1)
summary(res)
```

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```
  Q. Improve this plot by completing the below code, which adds color and axis labels


```{r}
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2)
mycols[ inds ] <- "blue"
plot(res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)")
```
  
```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res),  
                    keytype="ENSEMBL",  
                    column="SYMBOL",   
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),  
                    keytype="ENSEMBL",   
                    column="ENTREZID",   
                    multiVals="first")

res$name = mapIds(org.Hs.eg.db,
                  keys=row.names(res),   
                  keytype="ENSEMBL",   
                  column="GENENAME",  
                  multiVals="first")

head(res, 10)
```

Q. reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}
res = res[order(res$padj),]
write.csv(res, file="deseq_results.csv")
```

```{r}
library(pathview)
```

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```
  
  
```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```
  
```{r}
attributes(keggres)
```
  
```{r}
head(keggres$less)
```
  
  
```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```


```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```


```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```


 Q: What pathway has the most significant “Entities p-value”? Drug resistance of FLT3 mutants
 Do the most significant pathways listed match your previous KEGG results? No. 
 What factors could cause differences between the two methods? Different statistical models or databases used. 
 
 



















