---
title: "class 18"
format: html
editor: visual
---

Q1)

```{r}
options(repos = c(CRAN = "https://cran.rstudio.com"))
```

```{r}
install.packages("ggplot2")
```

```{r}
install.packages("datapasta")
```

```{r}
library(ggplot2)
library(datapasta)
```

```{r}
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L,2022L),
  No..Reported.Pertussis.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116,3044)
)
```

```{r}
head(cdc)
```

```{r}
ggplot(cdc) +
  aes(x = Year, y = No..Reported.Pertussis.Cases) +  # Adjust column names as needed
  geom_point() +  # Scatter plot
  geom_line() +  # Line plot
  labs(title = "Pertussis Cases by Year", x = "Year", y = "Number of Pertussis Cases")
```

Q2)

```{r}
ggplot(cdc) +
  aes(x = Year, y = No..Reported.Pertussis.Cases) +  # Plot pertussis cases by year
  geom_point() +  # Scatter plot
  geom_line() +  # Line plot
  geom_vline(xintercept = 1946, linetype = "dashed", color = "blue") +  # wP vaccine introduction
  geom_vline(xintercept = 1996, linetype = "dashed", color = "red") +   # aP vaccine switch
  labs(
    title = "Pertussis Cases by Year",
    x = "Year",
    y = "Number of Pertussis Cases"
  ) +
  theme_minimal()  # Optional: Adds a cleaner theme
```

Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

Pertussis cases increase again. In 2012 there were 48,277 cases in the US. This is the largest number of cases reported since 1955, where there were 62,786 cases.

This could be due to more sensitive PCR-based testing, vaccination hesitancy, or bacterial evolution.

```{r}
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
```

```{r}
head(subject, 3)
```

Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
# Count the number of Male and Female subjects
table(subject$biological_sex)
```

Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
today() - ymd("2000-01-01")
```

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```

Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
# wP
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

Q8. Determine the age of all individuals at time of boost?

```{r}
# Load necessary library
library(lubridate)

# Convert date_of_boost to Date and year_of_birth to a full date (e.g., January 1st of that year)
subject$date_of_boost <- ymd(subject$date_of_boost)  # Assuming it's already in 'YYYY-MM-DD' format
subject$year_of_birth <- ymd(paste(subject$year_of_birth, "01", "01", sep = "-"))  # Create a full date for year_of_birth

# Calculate the difference between boost date and birth date
int <- subject$date_of_boost - subject$year_of_birth

# Convert time difference to years
age_at_boost <- time_length(int, "year")

# Display the result
head(age_at_boost)
```

With the help of a faceted boxplot or histogram (below), do you think these two groups are significantly different? Yes they are statistacally different.

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

```{r}
install.packages("dplyr")
```

```{r}
install.packages("jsonlite")
library(jsonlite)
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```

9.  Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)
meta <- inner_join(subject,specimen)
head(meta)
```

Q10. using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

Q12. What are the different \$dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
#  the unique values of dataset column
unique(abdata$dataset)

#  the number of rows for each dataset
table(abdata$dataset)
```

Data collection decreased significantly after 2020.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

\*\*\*13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(x = MFI_normalised, y = antigen) +   # Use MFI_normalised for the x-axis
  geom_boxplot() +                          # Create a boxplot
  xlim(0, 75) +                             # Set the x-axis limits from 0 to 75
  facet_wrap(vars(visit), nrow = 2)         # Facet by 'visit' in 2 rows
```

Q14. What antigens show differences in IgG antibody titer levels t them over time? Why these and not others?

PRN, FHA, FIM 2/3, PT. Those are actually a part of the vaccine. The others are control.
